RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-22 doc/paper.tex
Perturbations are now finished.

    \begin{algorithm}
    \small
    \hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
    \hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
   	\begin{algorithmic}[1] 
   		\Procedure{RILS-ROLS}{$D_{tr}$}
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   		\EndProcedure
   	\end{algorithmic}
   	\caption{RILS-ROLS method.}
   	\label{alg:rilsrols}
   \end{algorithm}

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) around solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$ss \in s_{subtrees}$}
			\If{$ss = s$}
				\State $s_{perturbations} \gets s_{perturbations} \cup$ \texttt{Node1Perturbations}($ss$)
			\EndIf
			\If{$ss.arity \geq 1$}
				\State $node_{perturbations} \gets$ \texttt{Node1Perturbations}($ss.left$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.left$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
			\If{$ss.arity = 2$}
				\State $node_{perturbations} \gets$ \texttt{Node1Perturbations}($ss.right$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.right$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
		\EndFor
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations.}
	\label{alg:pert}
\end{algorithm}

>>>>

    \begin{algorithm}
    \small
    \hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
    \hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
   	\begin{algorithmic}[1] 
   		\Procedure{RILS-ROLS}{$D_{tr}$}
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{All1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   		\EndProcedure
   	\end{algorithmic}
   	\caption{RILS-ROLS method.}
   	\label{alg:rilsrols}
   \end{algorithm}

@@

Perturbations allow the algorithm to escape from local optima. As previously described, they are performed in two occasions: the first one during the exhaustive examination of neighboring solutions around the starting solution, and the second one being non-exhaustive, during selection of the next starting solution. 
In both cases, the same Algorithm~\ref{alg:pert} is used.

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) around solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$ss \in s_{subtrees}$} \label{line:forSSStart}
			\If{$ss = s$}
				\State $s_{perturbations} \gets s_{perturbations} \cup$ \texttt{NodeChange}($ss$)
			\EndIf
			\If{$ss.arity \geq 1$}
				\State $node_{perturbations} \gets$ \texttt{NodeChange}($ss.left$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.left$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
			\If{$ss.arity = 2$}
				\State $node_{perturbations} \gets$ \texttt{NodeChange}($ss.right$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.right$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations.}
	\label{alg:pert}
\end{algorithm}

Initially, the set of perturbations $s_{perturbations}$ is empty (line~\ref{line:pertInit}).

This is followed by constant normalization during which coefficients that enter multiplication, division, addition or subtraction are set to 1, while those entering the power function are rounded to integer, with exception of square root, which is kept intact. For example, for expression $3.3\cdot(x+45.1\cdot y^{3.2})\cdot 81\cdot x/\sqrt{y}$ the normalized version is $1\cdot (x+1\cdot y^3)\cdot 1\cdot x/\sqrt{y}$. The reason for performing normalization is to reduce the search space of possible perturbations. This reduction is reasonable, since normalization preserves the essential functional form. (Note that coefficients get tuned later during the OLS phase and local search.)

After normalization, the expression is simplified -- getting compact expression is more likely after normalization than before. The previous expression will take form $(x+y^3)\cdot x/\sqrt{y}$ -- this simplification only removed unnecessary coefficients, but in general it can be non-trivial symbolic simplification.

Perturbations are generated by making simple changes on the per-node level of $s$ expression tree.
Depending on the structure of the expression tree (expression does not have to have unique tree representation), the set of subtrees of the previous expression $(x+y^3)\cdot x/\sqrt{y}$ might be $\{(x+y^3)\cdot x/\sqrt{y}, (x+y^3), x/\sqrt{y}, x, y^3, \sqrt{y}, y\}$. This set of subtrees is obtained if left subtree of the whole expression is $(x+y^3)$ while the right one is $x/\sqrt{y}$. 
Lines \ref{line:forSSStart}-\ref{line:forSSEnd} show that there are three possibly overlapping cases when performing perturbations on the per-subtree level.

\emph{Case 1}. 
Subtree $ss$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be changed to addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Subtree $ss$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $ss=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is changed to division). 
Another example would be the case of unary operation, e.g. when $ss=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

\emph{Case 3}. 
Subtree $ss$ is binary operation, meaning the right subtree must exist. 
The idea is analog as in \emph{Case 2}.

Allowed per-node changes are as follows:

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate}
